---
title: "Thesis Code"
author: "Cade Stanley"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Models Using Pitch Data
First, I fit a logistic regression model and a random forest classifier that
include data about the pitch on which each stolen base attempt occurs. Knowing
the outcome of the pitch (including its speed, whether it is a pitchout, etc.)
may improve the predictive ability of the models, but this kind of information
will not be available to players and coaches when making a decision in game.

### Reading and Modifying the Data
```{r, results='hide', message=FALSE}
library(randomForest)
library(pROC)
library(caret)
set.seed(25)

full_data <- read.csv('full_data.csv', header=TRUE)

# change character data to uppercase
full_data <- data.frame(lapply(full_data, function(v) {
  if (is.character(v)) return(toupper(v))
  else return(v)
}))

# modify variable types
full_data$outcome <- as.factor(full_data$outcome)
full_data$on_third <- as.integer(as.logical(full_data$on_third))
full_data$is_hitters_count <- as.integer(as.logical(full_data$is_hitters_count))
full_data$is_pitchers_count <- as.integer(as.logical(full_data$is_pitchers_count))
full_data$strike_on_event <- as.integer(as.logical(full_data$strike_on_event))
full_data$swing_on_event <- as.integer(as.logical(full_data$swing_on_event))
full_data$pitchout_on_event <- as.integer(as.logical(full_data$pitchout_on_event))
full_data$blocked_on_event <- as.integer(as.logical(full_data$blocked_on_event))
full_data$pitcher_throws_left <- as.integer(as.logical(full_data$pitcher_throws_left))
full_data$batter_bats_left <- as.integer(as.logical(full_data$batter_bats_left))

# perform mean imputation
full_data$p_sb_rate_2017[is.na(full_data$p_sb_rate_2017)] <- mean(full_data$p_sb_rate_2017, na.rm=TRUE)
full_data$p_pickoff_rate_2018[is.na(full_data$p_pickoff_rate_2018)] <- mean(full_data$p_pickoff_rate_2018, na.rm=TRUE)
full_data$p_fastball_rate_2018[is.na(full_data$p_fastball_rate_2018)] <- mean(full_data$p_fastball_rate_2018, na.rm=TRUE)
full_data$r_sb_rate_2017[is.na(full_data$r_sb_rate_2017)] <- mean(full_data$r_sb_rate_2017, na.rm=TRUE)
full_data$r_sprint_speed_2018[is.na(full_data$r_sprint_speed_2018)] <- mean(full_data$r_sprint_speed_2018, na.rm=TRUE)
full_data$c_sb_rate_2017[is.na(full_data$c_sb_rate_2017)] <- mean(full_data$c_sb_rate_2017, na.rm=TRUE)
full_data$c_pop_2b_sba_2018[is.na(full_data$c_pop_2b_sba_2018)] <- mean(full_data$c_pop_2b_sba_2018, na.rm=TRUE)

# exclude observations that didn't occur on a pitch
full_data <- na.omit(full_data)
full_x <- full_data[,-1]
full_y <- full_data[,1]
full_k <- dim(full_data)[1]
```

### Logistic Regression
```{r, results='hide', message=FALSE}
# perform variable selection using a stepwise approach
full_stepwise_glm <- glm(outcome ~ outs + on_third + num_pitches + is_hitters_count + is_pitchers_count
                         + strike_on_event + swing_on_event + pitchout_on_event
                         + blocked_on_event + pickoffs_to_first + pitchouts
                         + pitches_run_on + b_score_difference + pitcher_throws_left
                         + batter_bats_left + start_speed + p_sb_rate_2017
                         + p_pickoff_rate_2018 + p_fastball_rate_2018 + r_sb_rate_2017
                         + r_sprint_speed_2018 + c_sb_rate_2017 + c_pop_2b_sba_2018
                         + p_sb_rate_2017*r_sb_rate_2017 + c_sb_rate_2017*r_sb_rate_2017
                         + c_pop_2b_sba_2018*r_sprint_speed_2018,
                         data=full_data, family="binomial")
best_full_glm <- step(full_stepwise_glm)
```
```{r}
summary(best_full_glm)
```

```{r, message=FALSE}
# restrict data to the best subset of variables
reduced_data <- subset(full_data, select=c(outcome, outs, on_third, num_pitches,
                                           is_hitters_count, is_pitchers_count,
                                           strike_on_event, pitchout_on_event, blocked_on_event,
                                           pitcher_throws_left, batter_bats_left, start_speed, 
                                           p_fastball_rate_2018, r_sb_rate_2017, r_sprint_speed_2018,
                                           c_pop_2b_sba_2018))
reduced_x <- reduced_data[,-1]
reduced_y <- reduced_data[,1]
reduced_k <- dim(reduced_data)[1]

# perform leave-one-out cross-validation on the best model
reduced_glm_predictions <- c()
for (i in 1:reduced_k) {
  reduced_glm <- glm(outcome ~ outs + on_third + num_pitches + is_hitters_count + is_pitchers_count
                     + strike_on_event + pitchout_on_event + blocked_on_event
                     + pitcher_throws_left + batter_bats_left + start_speed + p_fastball_rate_2018
                     + r_sb_rate_2017 + r_sprint_speed_2018 + c_pop_2b_sba_2018,
                     data=reduced_data[-i,], family="binomial")
  reduced_glm_predictions <- c(reduced_glm_predictions, predict(reduced_glm, newdata=reduced_x[i,], type="response"))
}

# calculate the AUC score and plot the ROC curve
reduced_glm_roc <- roc(reduced_y, reduced_glm_predictions)
reduced_glm_roc$auc
plot(reduced_glm_roc, main="Logistic Regression with Pitch Data")

# calculate the confusion matrix and related metrics
reduced_glm_class_predictions <- as.factor(ifelse(reduced_glm_predictions > 0.5, "SUCCESS", "FAILURE"))
confusionMatrix(data=reduced_glm_class_predictions, reference=reduced_y, positive="SUCCESS")
```

### Random Forest
```{r, message=FALSE}
# use 5-fold cross-validation (using more folds, or using the leave-one-out approach, is too expensive)
train_control <- trainControl(method = "cv", number = 5, savePredictions = TRUE, classProbs = TRUE)

full_rf <- train(outcome ~ outs + on_third + num_pitches + is_hitters_count + is_pitchers_count 
                 + strike_on_event + swing_on_event + pitchout_on_event
                 + blocked_on_event + pickoffs_to_first + pitchouts
                 + pitches_run_on + b_score_difference + pitcher_throws_left
                 + batter_bats_left + start_speed + p_sb_rate_2017
                 + p_pickoff_rate_2018 + p_fastball_rate_2018 + r_sb_rate_2017
                 + r_sprint_speed_2018 + c_sb_rate_2017 + c_pop_2b_sba_2018,
                 data=full_data, method="rf", trControl=train_control)

# calculate the AUC score and plot the ROC curve
full_rf_roc <- roc(full_rf$pred$obs, full_rf$pred$SUCCESS)
full_rf_roc$auc
plot(full_rf_roc, main="Random Forest with Pitch Data")

# calculate the confusion matrix and related metrics
confusionMatrix(data=full_rf$pred$pred, reference=full_rf$pred$obs, positive="SUCCESS")

# estimate variable importance
final_full_rf <- randomForest(outcome ~ outs + on_third + num_pitches + is_hitters_count + is_pitchers_count 
                              + strike_on_event + swing_on_event + pitchout_on_event
                              + blocked_on_event + pickoffs_to_first + pitchouts
                              + pitches_run_on + b_score_difference + pitcher_throws_left
                              + batter_bats_left + start_speed + p_sb_rate_2017
                              + p_pickoff_rate_2018 + p_fastball_rate_2018 + r_sb_rate_2017
                              + r_sprint_speed_2018 + c_sb_rate_2017 + c_pop_2b_sba_2018,
                              data=full_data, method="class", importance=TRUE)
importance(final_full_rf)
```

## Models Without Pitch Data
Now, I fit a logistic regression model and a random forest classifier without
using data about the pitches that occurred during a stolen base attempt. This
gives us around 200 more observations, as we can include stolen base attempts 
that did not occur on a pitch. Many of these observations are failed attempts, 
where the pitcher picks off the runner just as he is beginning to steal.

### Reading and Modifying the Data
```{r, message=FALSE}
pitchless_data <- read.csv('pitchless_data.csv', header=TRUE)

# change character data to uppercase
pitchless_data <- data.frame(lapply(pitchless_data, function(v) {
  if (is.character(v)) return(toupper(v))
  else return(v)
}))

# modify variable types
pitchless_data$outcome <- as.factor(pitchless_data$outcome)
pitchless_data$on_third <- as.integer(as.logical(pitchless_data$on_third))
pitchless_data$is_hitters_count <- as.integer(as.logical(pitchless_data$is_hitters_count))
pitchless_data$is_pitchers_count <- as.integer(as.logical(pitchless_data$is_pitchers_count))
pitchless_data$pitcher_throws_left <- as.integer(as.logical(pitchless_data$pitcher_throws_left))
pitchless_data$batter_bats_left <- as.integer(as.logical(pitchless_data$batter_bats_left))

# perform mean imputation
pitchless_data$p_sb_rate_2017[is.na(pitchless_data$p_sb_rate_2017)] <- mean(pitchless_data$p_sb_rate_2017, na.rm=TRUE)
pitchless_data$p_pickoff_rate_2018[is.na(pitchless_data$p_pickoff_rate_2018)] <- mean(pitchless_data$p_pickoff_rate_2018, na.rm=TRUE)
pitchless_data$p_fastball_rate_2018[is.na(pitchless_data$p_fastball_rate_2018)] <- mean(pitchless_data$p_fastball_rate_2018, na.rm=TRUE)
pitchless_data$r_sb_rate_2017[is.na(pitchless_data$r_sb_rate_2017)] <- mean(pitchless_data$r_sb_rate_2017, na.rm=TRUE)
pitchless_data$r_sprint_speed_2018[is.na(pitchless_data$r_sprint_speed_2018)] <- mean(pitchless_data$r_sprint_speed_2018, na.rm=TRUE)
pitchless_data$c_sb_rate_2017[is.na(pitchless_data$c_sb_rate_2017)] <- mean(pitchless_data$c_sb_rate_2017, na.rm=TRUE)
pitchless_data$c_pop_2b_sba_2018[is.na(pitchless_data$c_pop_2b_sba_2018)] <- mean(pitchless_data$c_pop_2b_sba_2018, na.rm=TRUE)

pitchless_x <- pitchless_data[,-1]
pitchless_y <- pitchless_data[,1]
pitchless_k <- dim(pitchless_data)[1]
```

### Logistic Regression
```{r, results='hide', message=FALSE}
# perform variable selection using a stepwise approach
pitchless_stepwise_glm <- glm(outcome ~ outs + on_third + num_pitches + is_hitters_count 
                              + is_pitchers_count + pickoffs_to_first + pitchouts 
                              + pitches_run_on + pitcher_throws_left + batter_bats_left
                              + p_sb_rate_2017 + p_pickoff_rate_2018 + p_fastball_rate_2018
                              + r_sb_rate_2017 + r_sprint_speed_2018 + c_sb_rate_2017
                              + c_pop_2b_sba_2018 + p_sb_rate_2017*r_sb_rate_2017 
                              + c_sb_rate_2017*r_sb_rate_2017 + c_pop_2b_sba_2018*r_sprint_speed_2018,
                              data=pitchless_data, family="binomial")
best_pitchless_glm <- step(pitchless_stepwise_glm)
```
```{r}
summary(best_pitchless_glm)
```

```{r, message=FALSE}
# restrict data to the best subset of variables
reduced_pitchless_data <- subset(pitchless_data, select=c(outcome, outs, on_third, num_pitches,
                                                          is_hitters_count, is_pitchers_count, pickoffs_to_first, 
                                                          pitchouts, pitches_run_on, r_sb_rate_2017, r_sprint_speed_2018,
                                                          c_pop_2b_sba_2018))
reduced_pitchless_x <- reduced_pitchless_data[,-1]
reduced_pitchless_y <- reduced_pitchless_data[,1]
reduced_pitchless_k <- dim(reduced_pitchless_data)[1]

# perform leave-one-out cross-validation on the best model
reduced_pitchless_glm_predictions <- c()
for (i in 1:reduced_pitchless_k) {
  reduced_pitchless_glm <- glm(outcome ~ outs + on_third + num_pitches + is_hitters_count 
                               + is_pitchers_count + pickoffs_to_first + pitchouts + pitches_run_on
                               + r_sb_rate_2017 + r_sprint_speed_2018 + c_pop_2b_sba_2018,
                               data=reduced_pitchless_data[-i,], family="binomial")
  reduced_pitchless_glm_predictions <- c(reduced_pitchless_glm_predictions, predict(reduced_pitchless_glm, 
                                                                                    newdata=reduced_pitchless_x[i,],
                                                                                    type="response"))
}

# calculate the AUC score and plot the ROC curve
reduced_pitchless_glm_roc <- roc(reduced_pitchless_y, reduced_pitchless_glm_predictions)
reduced_pitchless_glm_roc$auc
plot(reduced_pitchless_glm_roc, main="Logistic Regression without Pitch Data")

# calculate the confusion matrix and related metrics
reduced_pitchless_glm_class_predictions <- as.factor(ifelse(reduced_pitchless_glm_predictions > 0.5,
                                                            "SUCCESS", "FAILURE"))
confusionMatrix(data=reduced_pitchless_glm_class_predictions, reference=reduced_pitchless_y,
                positive="SUCCESS")
```

### Random Forest
```{r, message=FALSE}
pitchless_rf <- train(outcome ~ outs + on_third + num_pitches + is_hitters_count 
                      + is_pitchers_count + pickoffs_to_first + pitchouts 
                      + pitches_run_on + pitcher_throws_left + batter_bats_left
                      + p_sb_rate_2017 + p_pickoff_rate_2018 + p_fastball_rate_2018
                      + r_sb_rate_2017 + r_sprint_speed_2018 + c_sb_rate_2017 + c_pop_2b_sba_2018,
                      data=pitchless_data, method="rf", trControl=train_control)

# calculate the AUC score and plot the ROC curve
pitchless_rf_roc <- roc(pitchless_rf$pred$obs, pitchless_rf$pred$SUCCESS)
pitchless_rf_roc$auc
plot(pitchless_rf_roc, main="Random Forest without Pitch Data")

# calculate the confusion matrix and related metrics
confusionMatrix(data=pitchless_rf$pred$pred, reference=pitchless_rf$pred$obs, positive="SUCCESS")

# estimate variable importance
final_pitchless_rf <- randomForest(outcome ~ outs + on_third + num_pitches + is_hitters_count 
                                   + is_pitchers_count + pickoffs_to_first + pitchouts 
                                   + pitches_run_on + pitcher_throws_left + batter_bats_left
                                   + p_sb_rate_2017 + p_pickoff_rate_2018 + p_fastball_rate_2018
                                   + r_sb_rate_2017 + r_sprint_speed_2018 + c_sb_rate_2017 + c_pop_2b_sba_2018,
                                   data=pitchless_data, method="class", importance=TRUE)
importance(final_pitchless_rf)
```
